---
title: "Final Project 2"
author: "Ryan W."
date: "2024-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(dplyr)
library(readr)
library(lubridate)
library(tidyverse)
```


**Importing Data**

We first need to import our data by first reading in the csv files
```{r import_data, message=FALSE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv",
                "time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in,file_names)

global_cases <- read.csv(urls[1])
global_deaths <- read.csv(urls[2])
US_cases <- read.csv(urls[3])
US_deaths <- read.csv(urls[4])
```

**Tidying Up Data**

After looking at the data sets I would like to tidy up the data sets by narrowing down the number of variables into separate columns and renaming columns to be more R friendly.
I also need to remove the character "X" from data values in the date column and mutate the date to be more R friendly.
I also removed data values where cases were still at zero.

```{r tidy_global_data, message=FALSE}
global_cases <- global_cases %>% pivot_longer(cols = -c('Province.State', 
                'Country.Region', Lat, Long),
                names_to = "date", 
                values_to = "cases") %>% 
  select(-c(Lat, Long))

global_deaths <- global_deaths %>% pivot_longer(cols = -c('Province.State', 
                'Country.Region', Lat, Long),
                names_to = "date", 
                values_to = "deaths") %>% 
  select(-c(Lat, Long))

global_cases$date <- gsub("X", "", as.character(global_cases$date))
global_deaths$date <- gsub("X", "", as.character(global_deaths$date))

global <- global_cases %>% full_join(global_deaths) %>% 
  rename(Country_Region = 'Country.Region', 
         Province_State = 'Province.State') %>% 
  mutate(date = mdy(date))

global <- global %>% filter(cases > 0)
```

After tidying up the global data set, I also need to tidy up the US data set.

```{r tidy_US_data, message=FALSE}
US_cases <- US_cases %>% pivot_longer(cols = -(UID:Combined_Key), 
                                      names_to = "date", 
                                      values_to = "cases") %>% 
  select(Admin2:cases) %>% 
  select(-c(Lat, Long_))

US_deaths <- US_deaths %>% pivot_longer(cols = -(UID:Population), 
                                      names_to = "date", 
                                      values_to = "deaths") %>% 
  select(Admin2:deaths) %>% 
  select(-c(Lat, Long_))

US_cases$date <- gsub("X", "", as.character(US_cases$date))
US_deaths$date <- gsub("X", "", as.character(US_deaths$date))

US <- US_cases %>% full_join(US_deaths) %>% 
  mutate(date = mdy(date))
```

Since the US data set has a population variable which the global data set does not, I decided to add in a population column into the global data set.

```{r adding_pop}
global <- global %>% unite("Combined_Key", 
                           c(Province_State, Country_Region), 
                           sep = ", ", 
                           na.rm = TRUE, 
                           remove = FALSE)
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read.csv(uid_lookup_url) %>% 
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>% left_join(uid, by = c("Province_State", "Country_Region")) %>% 
  select(-c(UID, FIPS)) %>% 
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
```

**Visualizing and Analyzing Data**

First I need to break down the US data set further so it is arranged by states and then I take the totals of each state.

```{r more_US_data, message=FALSE}
US_by_state <- US %>% group_by(Province_State, Country_Region, date) %>% 
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% 
  mutate(deaths_per_mil = deaths * 1000000 / Population) %>% 
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mil, Population) %>% 
  ungroup()

US_totals <- US_by_state %>% group_by(Country_Region, date) %>% 
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% 
  mutate(deaths_per_mil = deaths * 1000000 / Population) %>% 
  select(Country_Region, date, cases, deaths, deaths_per_mil, Population) %>% 
  ungroup()
```

The graph below displays the total cases and deaths of COVID 19 throughout the pandemic in the US. As we can see at the beginning both cases and deaths skyrocketed but then toward the end it sort of plateaus.

```{r}
US_totals %>% filter(cases > 0) %>% 
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) + 
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() + 
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID19 in US", y = NULL)
```

The graph below displays the total cases and deaths of COVID 19 throughout the pandemic in California. As we can see that this graph is very similar to the graph above but California does not have as many cases and deaths. But that is to be expected as California is the US state with the highest population but California is a very large part of the US data set.

```{r}
state <- "California"
US_by_state %>% filter(Province_State == state) %>% filter(cases > 0) %>% 
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) + 
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() + 
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = str_c("COVID19 in ", state), y = NULL)
```

To get a better understanding of this data, I need create a new column that gives the number of new cases and new deaths for both US data sets.

```{r}
US_by_state <- US_by_state %>% 
  mutate(new_cases = cases - lag(cases), 
         new_deaths = deaths - lag(deaths))

US_totals <- US_totals %>% 
  mutate(new_cases = cases - lag(cases), 
         new_deaths = deaths - lag(deaths))
```

The graph below displays new cases and new deaths of COVID 19 throughout the pandemic in the US. We can see that the graph of new cases and new deaths is not linear and it constantly fluctuates between upward and downward trends. We can also see a large spike around the beginning of 2022 and toward the end of the graph the amount of new cases and deaths are on a steady downward trend.

```{r}
US_totals %>% ggplot(aes(x = date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) + 
  geom_point(aes(color = "new_cases")) + 
  geom_line(aes(y = new_deaths, color = "new_deaths")) + 
  geom_point(aes(y = new_deaths, color = "new_deaths")) + 
  scale_y_log10() + 
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID19 in US", y = NULL)
```

The graph below displays new cases and new deaths of COVID 19 throughout the pandemic in California. We can see that it is very similar to the graph above even the big spike at the beginning of 2022. One major difference between the two graphs is that the graph for California the number of new cases and deaths reach the bottom of the graph and we can assume that California has gotten control of COVID19.

```{r}
US_by_state %>% filter(Province_State == state) %>% 
  ggplot(aes(x = date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) + 
  geom_point(aes(color = "new_cases")) + 
  geom_line(aes(y = new_deaths, color = "new_deaths")) + 
  geom_point(aes(y = new_deaths, color = "new_deaths")) + 
  scale_y_log10() + 
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = str_c("COVID19 in ", state), y = NULL)
```

The first table displays the 10 US states with the lowest deaths per thousand. The second table displays the 10 US states with the highest deaths per thousand.

```{r}
US_state_totals <- US_by_state %>% group_by(Province_State) %>% 
  summarize(deaths = max(deaths), cases = max(cases), 
            population = max(Population), 
            cases_per_thou = 1000 * cases / population, 
            deaths_per_thou = 1000 * deaths / population) %>% 
  filter(cases > 0, population > 0)

US_state_totals %>% slice_min(deaths_per_thou, n = 10) %>% 
  select(deaths_per_thou, cases_per_thou, everything())

US_state_totals %>% slice_max(deaths_per_thou, n = 10) %>% 
  select(deaths_per_thou, cases_per_thou, everything())
```

The summary below is the summary of the prediction of deaths per thousand cases as a function of cases per thousand. This is a linear model as we will in the graph below.

```{r}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)
x_grid <- seq(1, 151)
new_df <- tibble(cases_per_thou = x_grid)
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))
```

This graph displays the recorded deaths per thousand as a function of cases per thousand in blue and the linear prediction model in red. As we can see the model roughly grasps the trend of the pandemic within the states. 

```{r}
US_tot_w_pred %>% ggplot() + 
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") + 
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```

**Conclusion**

So we conclude that COVID19 was uncontrolled throughout most of 2020 and had several spikes of new cases and new deaths at the beginning of 2021 and 2022. And by the tail of the data we can see the data plateau. Some US states handled COVID 19 very well by minimizing the number of deaths while other states did not handle COVID19 as well. Possible sources of bias are government withholding data or not accurately counting the number of cases and deaths, people not reporting that they have caught COVID19, and personal bias could have swayed the data. 
